Supplementary material for: non-coding RNAs of bird genomes

### 2014-03-11 paper/manuscript.tex
Amending manuscript, explaining losses, adding more expression data, fixing bugs

The ncRNAs pose serious research challenges, particularly for the
field of genomics. For example, they lack the strong statistical
signals associated with protein coding genes, e.g. open reading
frames, G+C content and codon-usage biases \cite{Rivas:2000}. Despite
promising techniques such as RNA-seq \cite{Croucher:2010}, homology
based methods, namely covariance models (CMs) remain state of the art
for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}.  The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

The CM based approach works well for almost all classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to
dramatic speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can
they deal simply with the repeats that many lncRNAs host. Consequently
in the latest release of Rfam the lncRNA families that were added were
composed of local conserved (and possibly structured elements) within
lncRNAs, analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. The functions determined to date for lncRNAs range
from regulating chromatin status to chromosomal inactivation
\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
genes is a lengthy and expensive process \cite{Guttman:2009}.

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
published 45 avian genomes \cite{birds:14,birds:14a}, provides
exciting opportunities to to explore ncRNA conservation in
unprecedented detail.

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. We focus our report on the unusual
results within the avian lineages. These are either unexpectedly
well-conserved RNAs or unexpectedly poorly-conserved RNAs. The former
are RNAs we would not have expected to be conserved between the birds
and the organisms these genes were initially identified in; Usually,
this hypothesis is based upon the function of the RNA which is not
conserved in avian species. The latter are apparent losses of RNA
genes that were expected to be conserved; usually, this hypothesis is
based upon the conservation of these RNAs in other vertebrate species.
We use three models of ``loss'' that explain the data: Firstly, these
could be genuine cases of an ancestral gene-loss along the avian
lineage. Secondly, this could be a case of ``divergence'' where a RNA
gene has undergone significant sequence and structural alterations, so
much so, that our homology search tools can no longer detect a
relationship between vertebrate exemplars and the avian
varieties. Thirdly, we consider the possibility that the available
genome assemblies have independently failed to capture these genes.

@@

In general, Rfam cannot include the entire length of any large, spliced
RNAs. This is a limitation of the covariance-models used for the
homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently, only
short, well-conserved regions with evolutionarily conserved secondary
structures are included in Rfam. By analogy to protein-domains, we
refer to these as RNA-domains \cite{Burge:2013}.

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus they
support a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
Figure~\ref{fig:5}).

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad ranch of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:5}).

Of the other classes of RNAs, none showed an unexpected degree of
conservation or expansion within the avian lineage. The only exception
being the snoRNA SNORD93 which we show has 92 copies in the tinamou
genome, whereas it only has 1-2 copies in all the other vertebrate
genomes.

\subsection*{RNA Losses, divergence or missing data?}

Much of the number of apparent losses and reduction in genomic
sequence has been extensively discussed elsewhere
\cite{Organ:2007}. Unsurprisingly, this reduction is reflected in the
copy-number of RNA genes. Some of the most dramatic examples are the
transfer RNAs and pseudogenes which average $\approx 900$ and $\approx
580$ copies in the human, turtle and alligator genomes, the average
copies numbers of these drop to $\approx 280$ and $\approx 100$ copies
in the avian genomes.

The absence of several well-conserved ncRNA families from many or even most
bird genomes is unlikely to represent true gene losses. This concerns in
particular the telomerase RNA, the RNA components of RNase P and MRP, the
minor spliceosomal snRNAs U4atac and U11, the selenocystein tRNA (tRNA-Sec)
as well as the vault RNAs. In order to get an idea to what extent the
absence of these RNAs from the \texttt{infernal-based} annotation is caused 
by sequence divergence beyond the thresholds of the Rfam CMs and/or 
missing or incomplete data, we complemented our analysis by dedicated
searches for a few of these RNA groups.

The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
for specificity and thus misses several occurrances that are easily
found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
sequences appear degraded at the ends, which may be explained e.g.\ by
low sequence quality at the very ends of contigs or scaffolds. A
\texttt{blastn} search also readily retrieves additional RNAse P and
RNAse MRP RNAs, capturing only the best conserved regions. In many
cases these additional candidates are incomplete or contain
undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Since \texttt{blastn}
searches remained unsuccessfull we constructed a sauropsid-specific CM
for the vault RNA. In addition to the hits identified by the Rfam
model we obtained three additional homologs. Vault RNAs, with a size
of about 100 nt, exhibit conserved sequence patterns only at their
ends, with essentially unconstrained sequence in the central
part. Their identification is one of the well-know difficult problems
for homology search \cite{Stadler:09b}.

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Figure~\ref{fig:7})
strongly suggests that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

@@

@@

Our diverse sets of genome annotations were combined. We collapsed the
overlapping annotations into a single annotation. We also generated
heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:5}
and Sup. Figs. 1-3). All the scripts and annotations presented here
are available from Github \cite{gitrepo}.

@@

@@

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/unusual-conserved.pdf}
  \includegraphics[width=0.45\textwidth]{figures/diverged.pdf}
  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    genomic copy number of ``unusually, well conserved RNAs''
    (particularly the lncRNAs) on the left and families that have been
    identified as surprising RNA Losses, divergence or missing
    data. In several cases functionally related families have also
    been included, e.g. the RNA components of the major and minor
    spliceosomes: U1, U2, U4, U5 and U6; and U11, U12, U4atac, U5 and
    U6atac, respectively.}\label{fig:5}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
  \caption[]{Additional homologs of some sparsely represented RNA
    families were discovered using dedicated search strategies
    combined with highly sensitive settings, synteny information,
    lineage-specific CMs and subsequent manual
    inspection.}\label{fig:7}
\end{figure}

>>>>

The ncRNAs pose serious research challenges, particularly for the
field of genomics. For example, they lack the strong statistical
signals associated with protein coding genes, e.g. open reading
frames, G+C content and codon-usage biases \cite{Rivas:2000}.

{\bf Say something about the promise of RNA-seq... but requires many
  tissues, developmental stages, fractions(e.g. poly-A, small, ...)
  and the FPR is unknown.}

Despite
promising techniques such as RNA-seq \cite{Croucher:2010}, homology
based methods, namely covariance models (CMs) remain state of the art
for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}.  The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

{\bf REDUCE?}
The CM based approach works well for almost all classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to
dramatic speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can
they deal simply with the repeats that many lncRNAs host. Consequently
in the latest release of Rfam the lncRNA families that were added were
composed of local conserved (and possibly structured elements) within
lncRNAs, analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. The functions determined to date for lncRNAs range
from regulating chromatin status to chromosomal inactivation
\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
genes is a lengthy and expensive process \cite{Guttman:2009}.

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
published 45 avian genomes \cite{birds:14,birds:14a}, provides
exciting opportunities to explore ncRNA conservation in
unprecedented detail. {\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
  CHROMOSOMES.}

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}.  {\bf Using accurate homology
  search tools and evolutionary constraints, we produce an accurate
  set of genome annotations for 48 bird species for ncRNAs that are
  conserved across the avian species. This conservative set of
  annotations is expected to contain the core avian ncRNAs.}
We focus our
report on the unusual results within the avian lineages. These are
either unexpectedly well-conserved RNAs or unexpectedly
poorly-conserved RNAs. The former are RNAs we would not have expected
to be conserved between the birds and the organisms these genes were
initially identified in; Usually, this hypothesis is based upon the
function of the RNA which is not conserved in avian species. The
latter are apparent losses of RNA genes that were expected to be
conserved; usually, this hypothesis is based upon the conservation of
these RNAs in other vertebrate species.  {\bf REWRITE: We use three
  models of ``loss'' that explain the data: Firstly, these could be
  genuine cases of an ancestral gene-loss along the avian
  lineage. Secondly, this could be a case of ``divergence'' where a
  RNA gene has undergone significant sequence and structural
  alterations, so much so, that our homology search tools can no
  longer detect a relationship between vertebrate exemplars and the
  avian varieties. Thirdly, we consider the possibility that the
  available genome assemblies have independently failed to capture
  these genes.}

@@

In general, Rfam cannot include the entire length of any large,
spliced RNAs. This is a limitation of the covariance-models used for
the homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently,
only short, well-conserved regions with evolutionarily conserved
secondary structures are included in Rfam. By analogy to
protein-domains, we refer to these as RNA-domains
\cite{Burge:2013}. {\bf SOME OF THESE MAY BE DNA ELEMENTS RATHER THAN
  LNCRNA ELEMENTS!!!!}

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus they
support a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
Figure~\ref{fig:1}).

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad ranch of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:1}).

Of the other classes of RNAs, none showed an unexpected degree of
conservation or expansion within the avian lineage. The only exception
being the snoRNA SNORD93 which has 92 copies in the tinamou
genome, whereas it only has 1-2 copies in all the other vertebrate
genomes.

\subsection*{RNA losses, divergence or missing data?}
Much of the number of apparent losses and reduction in genomic
sequence has been extensively discussed elsewhere
\cite{Organ:2007}. Unsurprisingly, this reduction is reflected in the
copy-number of RNA genes. Some of the most dramatic examples are the
transfer RNAs and pseudogenes which average $\sim900$ and $\sim580$
copies in the human, turtle and alligator genomes, the average copies
numbers of these drop to $\sim280$ and $\sim100$ copies in the avian
genomes.

The absence of seven well-conserved ncRNA families from many or even
most bird genomes is unlikely to represent true gene losses. This
concerns in particular the telomerase RNA, the RNA components of RNase
P and MRP, the minor spliceosomal snRNAs U4atac and U11, the
selenocystein tRNA (tRNA-Sec) as well as the vault RNAs (See
Figure~\ref{fig:1}).

{\bf Microchromosome background...}

We propose three possible models of ``loss'' to explain the data:
Firstly, these could be genuine cases of an ancestral gene-loss along
the avian lineage. Secondly, this could be a case of ``divergence''
where a RNA gene has undergone significant sequence and structural
alterations, so much so, that homology search tools no longer
detect a relationship between vertebrate exemplars and the avian
varieties. Thirdly, we consider the possibility that the available
genome assemblies have independently failed to capture these genes.

The third model of loss is the most likely explanation. These families
range in conservation level, from being ubiquitous to cellular-life
(RNase P RNA and tRNA-Sec), to present in the majority of eukaryotes
(RNase MRP, U4atac and U11) to present in most Bilateria
(Vault). Therefore, the loss or even diversification of these families
in the avian lineage is unlikely.

One defining feature of the avian genomes is the presence of
``microchromosomes''. As mentioned above, these are often difficult to
assemble. The most complete available avian genomes are the chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004} and the
zebrafinch \cite{Warren:2010}. The ``lost ncRNAs'' (loncRNAs) were
found to more prevalent in the more complete genomes 6/7 in chicken
and 3/7 in the zebrafinch. We examined the distribution of these
loncRNAs and discovered that only the chicken RNase MRP gene is found
on a macrochromosome. A Fisher's exact test showed that the loncRNAs
are significantly enriched within microchromosomes ($P<10^{16}$) in
both the chicken and the zebrafinch genomes (combined $P<2\times
10^{31}$).

{\bf CUT?}

In order to get an idea to what extent the
absence of these RNAs from the \texttt{infernal-based} annotation is
caused by sequence divergence beyond the thresholds of the Rfam CMs
and/or missing or incomplete data, we complemented our analysis by
dedicated searches for a few of these RNA groups.

The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
for specificity and thus misses several occurrances that are easily
found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
sequences appear degraded at the ends, which may be explained e.g.\ by
low sequence quality at the very ends of contigs or scaffolds. A
\texttt{blastn} search also readily retrieves additional RNase P and
RNAse MRP RNAs, capturing only the best conserved regions. In many
cases these additional candidates are incomplete or contain
undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Since \texttt{blastn}
searches remained unsuccessfull we constructed a sauropsid-specific CM
for the vault RNA. In addition to the hits identified by the Rfam
model we obtained three additional homologs. Vault RNAs, with a size
of about 100 nt, exhibit conserved sequence patterns only at their
ends, with essentially unconstrained sequence in the central
part. Their identification is one of the well-know difficult problems
for homology search \cite{Stadler:09b}.

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Figure~\ref{fig:2})
strongly suggests that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

@@

\section*{Pseudogenes}

{\bf Briefly mention the reduction in number of pseudogenes. Pick a
  few key human ones and compare with the birds. }

@@

Our diverse sets of genome annotations were combined. We collapsed the
overlapping annotations into a single annotation. We also generated
heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:1}
and Sup. Figs. 1-3). All the scripts and annotations presented here
are available from Github \cite{gitrepo}.

@@

Thanks to @ewanbirney for the following tweet: ``@ctitusbrown
@BioMickWatson So ... missing orthologs to chicken often mean 'gene
might be on the microchromosome'''.

We thank the anonymous referees for providing invaluable suggestions
that improved this work.

@@

\section*{Figures}
  \subsection*{Figure 1 - Heatmaps}

Heatmaps showing the prescence/abscence and approximate genomic copy
number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as
surprising RNA Losses, divergence or missing data. In several cases
functionally related families have also been included, e.g. the RNA
components of the major and minor spliceosomes: U1, U2, U4, U5 and U6;
and U11, U12, U4atac, U5 and U6atac, respectively.\label{fig:1}

%% \begin{figure}[ht]
%%   \includegraphics[width=0.45\textwidth]{figures/unusual-conserved.pdf}
%%   \includegraphics[width=0.45\textwidth]{figures/diverged.pdf}
%%   \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
%%   \caption[]{Heatmaps showing the prescence/abscence and approximate
%%     genomic copy number of ``unusually, well conserved RNAs''
%%     (particularly the lncRNAs) on the left and families that have been
%%     identified as surprising RNA Losses, divergence or missing
%%     data. In several cases functionally related families have also
%%     been included, e.g. the RNA components of the major and minor
%%     spliceosomes: U1, U2, U4, U5 and U6; and U11, U12, U4atac, U5 and
%%     U6atac, respectively.}\label{fig:1}
%% \end{figure}

  \subsection*{Figure 2 - Identifying missing RNAs}

Additional homologs of some sparsely represented RNA families were
discovered using dedicated search strategies combined with highly
sensitive settings, synteny information, lineage-specific CMs and
subsequent manual inspection.\label{fig:2}

{\bf TABLE ``AVERAGE'' NUMBER OF ncRNAs IN A BIRD GENOME?}

%% \begin{figure}[ht]
%%   \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
%%   \caption[]{Additional homologs of some sparsely represented RNA
%%     families were discovered using dedicated search strategies
%%     combined with highly sensitive settings, synteny information,
%%     lineage-specific CMs and subsequent manual
%%     inspection.}\label{fig:2}
%% \end{figure}

### 2014-03-11 paper/supplementary-material.tex
Amending manuscript, explaining losses, adding more expression data, fixing bugs

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/mir-17-cluster.pdf}
  \caption[Presence/absence table of mir-17 miRNA clusters]{This
    figure illustrates the distribution of all members of the mir-17
    clusters across birds, compared to fish and mammals. Columns
    correspond to the single miRNA sequences, ordered by cluster and
    position within its cluster. Rows correspond to the species and
    the cells contain the number of copies of the miRNA in the
    respective species. Colors correspond to orthologous miRNA
    families: mir-17 (orange), mir-19 (blue) and mir-92/25
    (green).There are 3 clusters, I-1 and I-X, and cluster II.  While
    the two copies of cluster I are completely conserved, cluster II
    has been lost as a whole in birds.}\label{fig:8}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/let-7-cluster.pdf}
  \caption[Presence/absence table of let-7 miRNA clusters]{This figure
    illustrates the distribution of the let-7 paralogs and clustered
    miRNAs of the families mir-100 and mir-125. Columns correspond to
    the single miRNA sequences, ordered by cluster and position within
    its cluster. Rows correspond to the species and the cells contain
    the number of copies of the miRNA in the respective
    species. Cluster A, which is strongly conserved in vertebrates has
    been completely lost in the avian lineage. Another obvious loss in
    birds is cluster F. However, in tinamou we detected an unclustered
    miRNA that best fits to F-let-7-1 in sequence and structure, while
    we could not find an I-let-7 miRNA sequence in this
    species.}\label{fig:9}
\end{figure}

>>>>

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/mir-17-cluster.pdf}
  \caption[Presence/absence table of mir-17 miRNA clusters]{
This figure illustrates the distribution of all members of the mir-17 clusters across birds, compared to fish and mammals. Columns correspond to single miRNA sequences, grouped by clusters and ordered by their position within the corresponding cluster. Rows correspond to the species and the cells contain the number of copies of the miRNA in the respective species. Colors correspond to orthologous miRNA families: mir-17 (orange), mir-19 (blue) and mir-92/25 (green).There are 3 clusters, I-1 and I-X, and cluster II.  While the two copies of cluster I are completely conserved, cluster II has been lost as a whole in birds.}\label{fig:8}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/let-7-cluster.pdf}
  \caption[Presence/absence table of let-7 miRNA clusters]{This figure illustrates the distribution of the let-7 paralogs and clustered miRNAs of the families mir-100 and mir-125. Columns correspond to single miRNA sequences, grouped by clusters and ordered by their position within the corresponding cluster. Rows correspond to the species and the cells contain the number of copies of the miRNA in the respective species. Cluster A, which is strongly conserved in vertebrates has been completely lost in the avian lineage. Another obvious loss in birds is cluster F. However, in tinamou we detected an unclustered miRNA that best fits to F-let-7-1 in sequence and structure, while we could not find an I-let-7 miRNA sequence in this species.}\label{fig:9}
\end{figure}
